Visualization and analysis of probability distributions of large temporal data

### 2021-11-04 thesis.Rmd
running without references

Placeholder

## Visualizing probability distributions across bivariate cyclic temporal granularities {#sec:gravitas}
## Detecting distributional differences between temporal granularities for exploratory time series analysis {#sec:hakear}
## Clustering time series based on probability distributions across temporal granularities {#sec:gracsr}
## Summary

@@

Placeholder

## Introduction
## Linear time granularities {#sec:linear-time}
### Definitions
### Relativities
## Cyclic time granularities {#sec:cyclic-gran}
### Circular granularities {#sec:circular-gran-def}
### Quasi-circular granularities {#sec:quasi-circular-gran-def}
### Aperiodic granularities {#sec:aperiodic-gran-def}
### Relativities {#sec:cyclic-calendar}
### Computation
#### Single-to-multiple order-up {-}
#### Multiple-to-single order-up {-}
#### Multiple order-up quasi-circular granularities {-}
## Data structure {#sec:data-structure}
### Harmonies and clashes {#sec:synergy}
### Near-clashes {#sec:near-clashes}
## Visualization {#sec:visualization}
### Data summarization
### Display choices for univariate distributions
### Comparison across sub-groups induced by conditioning {#sec:aesthetics}
#### Levels {-}
#### Synergy of cyclic granularities {-}
#### Conditioning variable {-}
## Applications {#sec:application}
### Smart meter data of Australia {#sec:smartmeter}
#### Cyclic granularities search and computation {-}
#### Screening and visualizing harmonies {-}
### T20 cricket data of Indian Premier League {#sec:cricket}
## Discussion {#sec:discussion}
## Acknowledgments {-}
## Supplementary Materials

>>>>

<!-- Introduction -->
<!-- Motivation & Background -->
<!-- Outline of the thesis -->

<!-- To tame the complexity of time, breaking it into years, months, weeks, days and so on in a hierarchical manner is a common way to relate data to time. Such dicrete human made abstractions of time can be thought of as time granularities.(Aigner et al., 2008). -->

<!-- introducing the overall field-->
With the availability of data at more and more finer time scales, exploration of time series data may be required to be carried out across both finer and coarser scales to draw useful inferences about the underlying process. For example, data collected at an hourly scale could be analyzed using coarser temporal scales such as days, months or quarters. This approach requires deconstructing time in various possible ways. Moreover, often it might be interesting to capture calendar or periodic effects like month-of-year, day-of-week or hour-of-day. They help us in answering questions like if certain levels of those time deconstructions are characterized by unusual/routine values of the observed variable. For example, certain days of the week or months of the year are likely to be characterized by higher values. It is important to be able to navigate through all the temporal deconstructions that accommodate for periodicities to have multiple perspectives of the observed data. This idea aligns with the notion of EDA (Tukey 1977) which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing.

<!-- outline the intro chapter structure -->

This chapter will provide an introduction to the study by first discussing the background and context, followed by the research aims, objectives and questions.

<!-- Background -->

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in household across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kwh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically, and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates of usage across households, total kwh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency.

<!-- introduce the specific problem  -->

<!-- ## change in data structure as a result of transitioning from linear to cyclic deconstructions -->
<!-- + Existing approaches, why we need to drill down, or why we want to -- probability distributions -->
However, restructuring time in this manner leads to restructured data where each level of the time deconstructions correspond to multiple values of the observed variable. It is common to see aggregation or summarization of these multiple observations with a unique value to study calendar effects. For example, using aggregates of usage across each hour/half-hour has been common in the literature because energy companies need to plan for maximum loads on the network. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency. Summarizing the probability distribution of these multiple observations to capture both the shape and uncertainty could be a potential way to understand the underlying distribution of these observations. Studying probability distributions is likely to focus on features of the data which are not transparent through raw data or a unique summary statistic.

<!-- Following research objectives would facilitate the achievement of this aim -->
<!-- Research objectives -->

<!-- ## Research aims -->

Hence, the overarching research goal is to study the periodic behavior of temporal data in a structured way by studying the probability distributions by best exploiting the characteristics of time. Slicing and dicing the data in all possible temporal scales as suggested by EDA can be a daunting task as it leads to a myriad of possibilities. This inspires the research presented in this thesis, which aims to provide a platform to systematically explore periodicities in temporal data and support finding regular patterns or anomalies, explore clusters of behaviors or summarize the behavior. The first part of the work discusses computation of all possible combinations of cyclic time granularities and a graphical mapping such that distributions of a numeric response variable is displayed across combinations of two cyclic granularities. Even analyzing the distribution of the measured variable across two cyclic granularities at once could amount to displaying many plots in search of potential  patterns. Thus, the first part of the research (@Gupta2020-vo) also introduces "harmony" to denote pairs of granularities that could be analyzed together and reduces the search from all possible options. But this approach is still overwhelming for human consumption because there would still be huge number of harmonies. Hence, the second part of the research extends this work and narrows the search further by finding pair of cyclic granularities which are informative enough and rank them according to their importance. However, to explore periodic patterns of many households, we have to resort to clustering which has been addressed in the third part of the research. Although the motivation came through the smart meter example, this is a problem that is relevant to any temporal data observed more than once per year.

## Visualizing probability distributions across bivariate cyclic temporal granularities {#sec:gravitas}

Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes classes of time deconstructions using linear and cyclic time granularities. Linear granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic granularities can be circular such as hour-of-the-day, quasi-circular such as day-of-the-month, and aperiodic such as public holidays. The hierarchical structure of granularities creates a nested ordering: hour-of-the-day and second-of-the-minute are single-order-up. Hour-of-the-week is multiple-order-up, because it passes over day-of-the-week. Methods are provided for creating all possible granularities for a time index. A recommendation algorithm provides an indication whether a pair of granularities can be meaningfully examined together (a "harmony"), or when they cannot (a "clash").

 Time granularities can be used to create data visualizations to explore for periodicities, associations and anomalies. The granularities form categorical variables (ordered or unordered) which induce groupings of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables.

 The methods implemented in the open source R package `gravitas` are consistent with a tidy workflow, with probability distributions examined using the range of graphics available in `ggplot2`.

## Detecting distributional differences between temporal granularities for exploratory time series analysis {#sec:hakear}

Cyclic temporal granularities, which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. The granularities feed new approaches to exploring time series data. One use is to take pairs of granularities, and make plots of response values across the categories induced by the temporal deconstruction. However, when there are many granularities that can be constructed for a time period, there will also be too many possible displays to decide which might be the more interesting to display. This work proposes a new distance metric to screen and rank the possible granularities, and hence choose the most interesting ones to plot. The distance measure is computed for a single or pairs of cyclic granularities can can be compared across different cyclic granularities and also on a collection of time series. The methods are implemented in the open-source R package `hakear`.

## Clustering time series based on probability distributions across temporal granularities {#sec:gracsr}

 With more and more time series data being collected at much finer temporal resolution, for a longer length of time, and for a larger number of individuals/entities, time series clustering research is getting a lot of traction. The sort of noisy, patchy, uneven, and asynchronous time series that is typical in many disciplines limits similarity searches among these lengthy time series. In this work, we suggest a method for overcoming these constraints by grouping time series based on probability distributions over cyclic temporal granularities. Cyclic granularities are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, and so on, and can be helpful for detecting repeating patterns. Looking at probability distributions across cyclic granularities results in an approach that is robust to missing or noisy data, aids in dimension reduction, and ensures small pockets of similar repeated behaviours. The proposed method was tested using a collection of residential electricity customers. The simulated and empirical evidence demonstrates that our method is capable of producing meaningful clusters.

<!-- Time series clustering research is gaining a lot of importance with more and more time series data recorded at much finer temporal resolution, for a longer period of time and for a larger number of individuals/entities.Similarity searches amongst these long time series are often limited by the type of noisy, patchy, unequal and asynchronous time series common in many fields. In this paper we propose a strategy to alleviate these limitations by  clustering time series based on probability distributions across cyclic temporal granularities. Cyclic granularities are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, which can be useful for measuring repetitive patterns in large univariate time series data. Thus, looking at the probability distributions across cyclic granularity leads to a method which is robust to missing or noisy data, helps in dimension reduction while ensuring small pockets of similar repetitive behaviors. The method is applied to electricity smart meter dat -->
<!-- The suggested approach was evaluated on a set of benchmark time series on residential electricity customers. The empirical results show that our approach is able to yield meaningful clusters. -->

## Summary

The thesis is structured as follows. Chapter \@ref(ch:gravitas) provides details of the cyclic granularities, different classes, and computation, and also its usage in exploratory time series analysis through applications. This is implemented in the R package **gravitas**. Chapter \@ref(ch:hakear) provides guidance on how to choose significant cyclic granularities, which are likely to have interesting patterns across its categories. This is available as the R package **hakear**. The chapter @ref (ch: gracsr) provides similarity measures for comparing multiple time series.This is in the developing R package **gracsr**. Chapter \@ref(ch:conclusion) summarizes the software tools developed for the work, and discusses some future plans.

@@

Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes classes of time deconstructions using linear and cyclic time granularities. Linear granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic granularities can be circular such as hour-of-the-day, quasi-circular such as day-of-the-month, and aperiodic such as public holidays. The hierarchical structure of granularities creates a nested ordering: hour-of-the-day and second-of-the-minute are single-order-up. Hour-of-the-week is multiple-order-up, because it passes over day-of-the-week. Methods are provided for creating all possible granularities for a time index. A recommendation algorithm provides an indication whether a pair of granularities can be meaningfully examined together (a "harmony"), or when they cannot (a "clash").

 Time granularities can be used to create data visualizations to explore for periodicities, associations and anomalies. The granularities form categorical variables (ordered or unordered) which induce groupings of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables.

 The methods implemented in the open source R package `gravitas` are consistent with a tidy workflow, with probability distributions examined using the range of graphics available in `ggplot2`.

```{r external, include = FALSE, cache = FALSE}
read_chunk('scripts/gravitas.R')
```

```{r load}
```

## Introduction

Temporal data are available at various resolutions depending on the context. Social and economic data are often collected and reported at coarse temporal scales such as monthly, quarterly or annually. With recent advancements in technology, more and more data are recorded at much finer temporal scales. Energy consumption may be collected every half an hour, energy supply may be collected every minute, and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales such as days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities [@aigner2011visualization].

It is important to be able to navigate through all of these time granularities to have multiple perspectives on the periodicity of the observed data. This aligns with the notion of exploratory data analysis (EDA) [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is an indispensable tool for exploration. Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task overwhelming.

Calendar-based graphics [@wang2020calendar] are useful in visualizing patterns in the weekly and monthly structure and are helpful when checking for the effects of weekends or special days. Any temporal data at sub-daily resolution can also be displayed using this type of faceting [@Wickham2009pk] with days of the week, month of the year, or another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions. There can be many different time deconstructions, based on the calendar or on categorizations of time granularities.

Linear time granularities (such as hours, days, weeks and months) respect the linear progression of time and are non-repeating. One of the first attempts to characterize these granularities is due to @Bettini1998-ed. However, the definitions and rules defined are inadequate for describing non-linear granularities. Hence, there is a need to define some new time granularities, that can be useful in visualizations. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an example of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities such as second of the minute, or multiple-order-up granularities such as second of the hour. The lubridate package [@Grolemund2011-vm] provides tools to access and manipulate common date-time objects. However, most of its accessor functions are limited to single-order-up granularities.

The motivation for this work stems from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties as well as physical properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially; for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates [see @Goodwin_2012] of usage across households, such as half-hourly total usage by state, because energy companies need to plan for maximum loads on the network. But studying overall energy use hides the distribution of usage at finer scales, and makes it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data will benefit from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns and anomalies. Although we were motivated by the smart meter example, the problem and the solutions we propose are practically relevant to any temporal data observed more than once per year. In a broader sense, it could be even suitable for data observed by years, decades, and centuries as might be in weather or astronomical data.

This work provides tools for systematically exploring bivariate granularities within the tidy workflow (@wickham2016r). In particular, we

  * provide a formal characterization of cyclic granularities;
  * facilitate manipulation of single- and multiple-order-up time granularities through cyclic calendar algebra;
  * develop an approach to check the feasibility of creating plots or drawing inferences for any two cyclic granularities.

The remainder of the paper is organized as follows: Section&nbsp;\ref{sec:linear-time} provides some background material on linear granularities and calendar algebra for computing different linear granularities. Section&nbsp;\ref{sec:cyclic-gran} formally characterizes different cyclic time granularities by extending the framework of linear time granularities, and introducing cyclic calendar algebra for computing cyclic time granularities. The data structure for exploring the conditional distributions of the associated time series across pairs of cyclic time granularities is discussed in Section&nbsp;\ref{sec:data-structure}. Section&nbsp;\ref{sec:visualization} discusses the role of different factors in constructing an informative and trustworthy visualization. Section&nbsp;\ref{sec:application} examines how systematic exploration can be carried out for a temporal and non-temporal application. Finally, we summarize our results and discuss possible future directions in Section&nbsp;\ref{sec:discussion}.

## Linear time granularities {#sec:linear-time}

Discrete abstractions of time such as weeks, months or holidays can be thought of as "time granularities". Time granularities are **linear** if they respect the linear progression of time. There have been several attempts to provide a framework for formally characterizing time granularities, including @Bettini1998-ed which forms the basis of the work described here.

### Definitions

\begin{defn}\label{def:definition}
A \textbf{time domain} is a pair $(T; \le)$ where $T$ is a non-empty set of time instants (equivalently, moments or points) and $\le$ is a total order on $T$.
\end{defn}

\noindent The time domain is assumed to be *discrete*, and there is unique predecessor and successor for every element in the time domain except for the first and last.

\begin{defn}\label{def:index set}
The \textbf{index set}, $Z=\{z: z \in \mathbb{Z}_{\geq 0}\}$, uniquely maps the time instants to the set of non-negative integers.
\end{defn}

\begin{defn}\label{def:linear}
A \textbf{linear granularity} is a mapping $G$ from the index set, $Z$, to subsets of the time domain such that:
  (1) if $i < j$ and $G(i)$ and $G(j)$ are non-empty, then each element of $G(i)$ is less than all elements of $G(j)$; and
  (2) if $i < k < j$ and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty.
Each non-empty subset $G(i)$ is called a \textbf{granule}.
\end{defn}

\noindent This implies that the granules in a linear granularity are non-overlapping, continuous and ordered. The indexing for each granule can also be associated with a textual representation, called the label. A discrete time model often uses a fixed smallest linear granularity named by @Bettini1998-ed **bottom granularity**. \autoref{fig:linear-time} illustrates some common linear time granularities. Here, "hour" is the bottom granularity and "day", "week", "month" and "year" are linear granularities formed by mapping the index set to subsets of the hourly time domain. If we have "hour" running from $\{0, 1, \dots,t\}$, we will have "day" running from $\{0, 1, \dots, \lfloor t/24\rfloor\}$. These linear granularities are uni-directional and non-repeating.

```{r linear-time, echo=FALSE, fig.cap="(ref:linear-time)", fig.pos="!htbp"}
```

(ref:linear-time) Illustration of time domain, linear granularities and index set. Hour, day, week, month and year are linear granularities and can also be considered to be time domains. These are ordered with ordering guided by integers and hence is unidirectional and non-repeating. Hours could also be considered the index set, and a bottom granularity.

### Relativities

Properties of pairs of granularities fall into various categories.

\begin{defn}\label{def:finerthan}
A linear granularity $G$ is \textbf{finer than} a linear granularity $H$, denoted $G \preceq H$, if for each index $i$, there exists an index $j$ such that
$G(i) \subset H(j).$
\end{defn}

\begin{defn}\label{def:groupsinto}
A linear granularity $G$ \textbf{groups into} a linear granularity $H$, denoted
$G \trianglelefteq H$, if for each index $j$ there exists a (possibly infinite) subset $S$ of the integers such that $H(j) = \bigcup_{i \in S}G(i).$
\end{defn}

\noindent For example, both $day \trianglelefteq week$ and $day \preceq week$ hold, since every granule of $week$ is the union of some set of granules of day and each day is a subset of a $week$. These definitions are not equivalent. Consider another example, where $G_1$ denotes "weekend" and $H_1$ denotes "week". Then, $G_1 \preceq H_1$, but $G_1 \ntrianglelefteq H_1$. Further, with $G_2$ denoting "days" and $H_2$ denoting "business-week", $G_2 \npreceq H_2$, but $G_2 \trianglelefteq H_2$, since each business-week can be expressed as an union of some days, but Saturdays and Sundays are not subset of any business-week.
Moreover, with  $H_3$ denoting "public holidays", $G_1 \npreceq H_3$ and $G_1 \ntrianglelefteq H_3$.

<!-- # talk about period after definition 6 after you have explained period -->
<!-- The relationship has period 7. -->
<!-- The relationship $day \trianglelefteq month$ has a more complicated period. If leap years are ignored, each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year. With the inclusion of leap years, the grouping period is 400 years. -->

\begin{defn}\label{def:periodic}
A granularity $G$ is \textbf{periodic} with respect to a finite granularity $H$ if:
(1) $G \trianglelefteq H$; and
(2) there exist $R$, $P \in \mathbb{Z}_+$, where $R$ is less than the number of granules of $H$, such that for all $i \in \mathbb{Z}_{\ge {0}}$ , if $H(i) = \bigcup_{j \in S}G(j)$ and $H (i + R) \neq \phi$ then $H (i + R) = \bigcup_{j \in S} G(j + P)$.
\end{defn}

If $G$ groups into $H$, it would imply that any granule $H(i)$ is the union of some granules of $G$, for example, $G(a_1), G(a_2), \dots, G(a_k)$. Condition (2) in Definition \ref{def:periodic} implies that if $H(i + R) \neq \emptyset$, then $H(i + R) = \bigcup (G(a_1 + P), G(a_2 + P), \dots, G(a_k + P))$, resulting in a "periodic" pattern of the composition of $H$ using granules of $G$. In this pattern, each granule of $H$ is shifted by $P$ granules of $G$. $P$ is called the \textbf{Period} (@Bettini2000-qk).

For example, day is periodic with respect to week with $R=1$ and $P=7$, while (if we ignore leap years) day is periodic with respect to month with $R=12$ and $P=365$ as any month would consist of the same number of days across years.
Since the idea of period involves a pair of granularities, we say that the pair $(day, week)$ has period 7, while the pair $(day, month)$ has a period 365 (ignoring leap years).

Granularities can also be periodic with respect to other granularities, _"except for a finite number of spans of time where they behave in an anomalous way"_; these are called **quasi-periodic** relationships [@Bettini2000-vy]. In a Gregorian calendar with leap years, day groups quasi-periodically into month with the exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.

\begin{defn}\label{def:order}
The \textbf{order} of a linear granularity is the level of coarseness associated with a linear granularity. A linear granularity G will have lower order than H if each granule of G is composed of lower number of granules of bottom granularity than each granule of H.
\end{defn}

With two linear granularities $G$ and $H$, if $G$ *groups into* or *finer than* $H$ then $G$ is of lower order than $H$. For example, if the bottom granularity is hour, then granularity $day$ will have lower order than $week$ since each day consist of fewer hours than each week.

Granules in any granularity may be aggregated to form a coarser granularity. A system of multiple granularities in lattice structures is referred to as a **calendar** by @Dyreson_2000. Linear time granularities are computed through "calendar algebra" operations [@Ning_2002] designed to generate new granularities recursively from the bottom granularity. For example, due to the constant length of day and week, we can derive them from hour using
$$
  D(j) = \lfloor H(i)/24\rfloor, \qquad W(k) = \lfloor H(i)/(24*7)\rfloor,
$$
where $H$, $D$ and $W$ denote hours, days and weeks respectively.

## Cyclic time granularities {#sec:cyclic-gran}

Cyclic granularities represent cyclical repetitions in time. They can be thought of as additional categorizations of time that are not linear. Cyclic granularities can be constructed from two linear granularities, that relate periodically; the resulting cycles can be either _regular_ (**circular**), or _irregular_ (**quasi-circular**).

### Circular granularities {#sec:circular-gran-def}

\begin{defn}\label{def:circular}
A \textbf{circular granularity} $C_{B, G}$ relates linear granularity $G$ to bottom granularity $B$ if
\begin{equation} \label{eq:circular-gran}
\begin{split}
C_{B, G}(z) & = z\mod P(B, G) \quad \forall z \in \mathbb{Z}_{\geq 0} \\
\end{split}
\end{equation}
where
$z$ denotes the index set,
$B$ groups periodically into $G$ with regular mapping and period $P(B, G)$.
\end{defn}

```{r circular-dow, echo=FALSE, fig.cap= "(ref:circular-dow)",fig.pos="!htb"}
```

(ref:circular-dow) Index sets for some linear and circular granularities (a). Circular granularities can be constructed by slicing the linear granularity into pieces and stacking them (b).

\autoref{fig:circular-dow} illustrates some linear and cyclical granularities. Cyclical granularities are constructed by cutting the linear granularity into pieces, and stacking them to match the cycles (as shown in b). $B, G, H$ (day, week, fortnight, respectively) are linear granularities. The circular granularity $C_{B, G}$ (day-of-week) is constructed from $B$ and $G$, while  circular granularity $C_{B, H}$ (day-of-fortnight) is constructed from $B$ and $H$. These overlapping cyclical granularities share elements from the linear granularity. Each of $C_{B , G}$ and $C_{B , H}$ consist of repeated patterns $\{0, 1, \dots, 6\}$ and $\{0, 1, \dots, 13\}$ with $P=7$ and $P=14$ respectively.

Suppose ${L}$ is a label mapping that defines a unique label for each index $\ell \in \{ 0,1,\dots, (P-1)\}$. For example, the label mapping $L$ for $C_{B, G}$ can be defined as
$$
  L: \{0,1, \dots, 6\} \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}.
$$

In general, any circular granularity relating two linear granularities can be expressed as
$$
  C_{G, H}(z) = \lfloor z/P(B,G) \rfloor\mod P(G,H),
$$
where $H$ is periodic with respect to $G$ with regular mapping and period $P(G,H)$. Table&nbsp;\ref{tab:definitions} shows several circular granularities constructed using minutes as the bottom granularity.

\begin{table}[ht]
\begin{center}
\begin{tabular}{lll}
\toprule
Circular granularity & Expression & Period \\
\midrule
minute-of-hour                               &
  $C_1 = z \mod 60$                     &
  $P_1 = \phantom{99}60$ \\
minute-of-day                                &
  $C_2 = z \mod 60*24$                  &
  $P_2= `r 60*24`$\\
hour-of-day                                  &
  $C_3 = \lfloor z/60\rfloor\mod 24$    &
  $P_3 = \phantom{99}24$ \\
hour-of-week                                 &
  $C_4 = \lfloor z/60\rfloor\mod 24*7$  &
  $P_4= \phantom{9}`r 24*7`$\\
day-of-week                                  &
  $C_5 = \lfloor z/24*60\rfloor \mod 7$ &
  $P_5= \phantom{999}7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and period $P_i$. Circular granularities can be expressed using modular arithmetic due to their regular mapping.}
\label{tab:definitions}
\end{table}

### Quasi-circular granularities {#sec:quasi-circular-gran-def}

A **quasi-circular** granularity cannot be defined using modular arithmetic because of the irregular mapping. However, they are still formed with linear granularities, one of which groups periodically into the other. \autoref{tab:quasi} shows some examples of quasi-circular granularities.

\begin{table}[ht]
\centering
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
Quasi-circular granularity && Possible period lengths\\
\midrule
$Q_1 =$ day-of-month && $P_1 = 31, 30, 29, 28$\\
$Q_2 =$ hour-of-month && $P_2 = 24\times 31, 24\times 30, 24\times 29, 24\times 28$\\
$Q_3 =$ day-of-year && $P_3 = 366, 365$\\
\bottomrule
\end{tabular}
\caption{Examples of quasi-circular granularities relating two linear granularities with irregular mapping leading to several possible period lengths.}
\label{tab:quasi}
\end{table}

\begin{defn}\label{def:quasicircular}
A \textbf{quasi-circular granularity} $Q_{B, G'}$ is formed when bottom granularity $B$ groups periodically into linear granularity $G'$ with irregular mapping such that the granularities are given by
\begin{equation}\label{eq:quasi}
Q_{B, G'}(z) =
z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert, \quad \text{for}\quad z \in T_{k},
\end{equation}
where
$z$ denotes the index set,
$w$ denotes the index of $G'$,
$R'$ is the number of granules of $G'$ in each period of $(B, G')$,
$T_w$ are the sets of indices of $B$ such that $G'(w) = \bigcup_{z \in T_w}B(z)$,
and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{defn}

For example, day-of-year is quasi-periodic with either 365 or 366 granules of $B$ (days) within each granule of $G'$ (years). The pattern repeats every $4$ years (ignoring leap seconds). Hence $R'= 4$. $Q_{B, G'}$ is a repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not the same across different granules of $G'$.

### Aperiodic granularities {#sec:aperiodic-gran-def}

Aperiodic linear granularities are those that cannot be specified as a periodic repetition of a pattern of granules as described in Definition \ref{def:periodic}. Aperiodic cyclic granularities capture repetitions of these aperiodic linear granularities.
Examples include public holidays which repeat every year, but there is no reasonably small span of time within which their behavior remains constant. A classic example is Easter (in the western tradition) whose dates repeat only after 5.7 million years [@Reingold2001-kf]. In Australia, if a standard public holiday falls on a weekend, a substitute public holiday will sometimes be observed on the first non-weekend day (usually Monday) after the weekend. Examples of aperiodic granularity may also include school holidays or a scheduled event. All of these are recurring events, but with non-periodic patterns. Consequently, $P_i$ (as given in \autoref{tab:quasi}) are essentially infinite for aperiodic granularities.

\begin{defn}\label{def:aperiodic}
An \textbf{aperiodic cyclic granularity} is formed when bottom granularity $B$ groups into an aperiodic linear granularity $M$ such that the granularities are given by
\begin{equation}\label{eq:aperiodic}
A_{B, M}(z) = \begin{cases}
                  i, & \text{for}\quad z \in T_{i_j} \\
                  0  & \text{otherwise},
                \end{cases}
\end{equation}
where
$z$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, and $M = \bigcup_{i=1}^{n}M_{i}$, $n$ being the number of  aperiodic linear granularities in consideration.
\end{defn}

```{r aperiodic-example, echo=FALSE, fig.cap= "(ref:aperiodic-example)", fig.pos="!htb"}
```

(ref:aperiodic-example) Quasi-circular and aperiodic cyclic granularities illustrated by linear (a) and 
stacked-displays (b) progression of time. The linear display shows the granularities days ($B$), weeks ($G$), semester days ($B'$), and stages of a semester ($M$) indexed over a linear representation of time. The granules of $B'$ is only defined for days when the semester is running. Here a semester spans 18 weeks and 2 days, and consists of 6 stages. It starts with a week of orientation, followed by an in-session period (6 weeks), a break (1 week), the second half of semester (7 weeks), a 1-week study break before final exams, which spans the next 16 days. This distribution of semester days remains relatively similar for every semester. $Q_{B',M}$ with P = 128 is a quasi-circular granularity with repeating patterns, while $A_{B,M}$ is an aperiodic cyclic granularity as the placement of the semester within a year varies from year to year with no fixed start and end dates.

For example, consider the school semester shown in \autoref{fig:aperiodic-example}. Let the linear granularities $M_1$ and $M_2$ denote the teaching and non-teaching stages of the semester respectively. Both $M_1$, $M_2$ and $M = M_{1}\bigcup M_{2}$ denoting the "stages" of the semester are aperiodic with respect to days ($B$) or weeks ($G$). Hence $A_{B, M}$ denoting day-of-the-stage would be an aperiodic cyclic granularity because the placement of the semester within a year would vary across years. Here, $Q_{B', M}$ denoting semester-day-of-the-stage would be a quasi-circular granularity
since the distribution of semester days within a semester is assumed to remain constant over years. Here semester-day is denoted by "sem day" ($B'$) and its granules are only defined for the span of the semesters.

### Relativities {#sec:cyclic-calendar}

The hierarchical structure of time creates a natural nested ordering which can be used in the computation of relative pairs of granularities.

\begin{defn}\label{def:hierarchy}
The nested ordering of linear granularities can be organized into a \textbf{hierarchy table}, denoted as $H_n: (G, C, K)$, which arranges them from lowest to highest in order. It shows how the $n$ granularities relate through $K$, and how the cyclic granularities, $C$, can be defined relative to the linear granularities. Let $G_{\ell}$ and $G_{m}$ represent the linear granularity of order $\ell$ and $m$ respectively with $\ell<m$. Then $K \equiv P(\ell,m)$ represents the period length of the grouping $(G_{\ell}, G_{m})$, if $C_{G_{\ell}, G_{m}}$ is a circular granularity and $K \equiv k(\ell,m)$ represents the operation to obtain $G_{m}$ from $G_{\ell}$, if $C_{G_{\ell}, G_{m}}$ is quasi-circular.
\end{defn}

For example, \autoref{tab:tab-mayan} shows the hierarchy table for the Mayan calendar. In the Mayan calendar, one day was referred to as a kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal (about a year); 1 katun = 20 tun (20 years) and 1 baktun = 20 katun.

```{r tab-mayan}
```

Like most calendars, the Mayan calendar used the day as the basic unit of time [@Reingold2001-kf]. The structuring of larger units, weeks, months, years and cycle of years, though, varies substantially between calendars. For example,  the French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds", the duration of which is 0.864 common seconds. Nevertheless, for any calendar, a hierarchy table can be defined. Note that it is not always possible to organize an aperiodic linear granularity in a hierarchy table. Hence, we assume that the hierarchy table consists of periodic linear granularities only, and that the cyclic granularity $C_{G(\ell),G(m)}$ is either circular or quasi-circular.

\begin{defn}\label{def:norderup}
The hierarchy table contains \textbf{multiple-order-up} granularities which are cyclic granularities that are nested within multiple levels.
A \textbf{single-order-up} is a cyclic granularity which is nested within a single level. It is a special case of multiple-order-up granularity.
\end{defn}

\noindent In the Mayan calendar (Table \@ref(tab:tab-mayan)), kin-of-tun or kin-of-baktun are examples of multiple-order-up granularities and single-order-up granularities are kin-of-uinal, uinal-of-tun etc.

### Computation

Following the calendar algebra of @Ning_2002 for linear granularities, we can define cyclic calendar algebra to compute cyclic granularities. Cyclic calendar algebra comprises two kinds of operations:
(1) **single-to-multiple** (the calculation of _multiple-order-up_ cyclic granularities from _single-order-up_ cyclic granularities) and (2) **multiple-to-single** (the reverse).

#### Single-to-multiple order-up {-}

Methods to obtain multiple-order-up granularity will depend on whether the hierarchy consists of all circular single-order-up granularities or a mix of circular and quasi-circular single-order-up granularities. Circular single-order-up granularities can be used recursively to obtain a multiple-order-up circular granularity using
\begin{equation} \label{eq:eq7}
C_{G_\ell,G_m}(z)
  = \sum_{i=0}^{m - \ell - 1} P(\ell, \ell+i)C_{G_{\ell+i},G_{\ell+i+1}}(z),
\end{equation}
where $\ell < m - 1$ and $P(i, i) = 1$ for $i=0,1,\dots,m-\ell-1$, and
$C_{B, G}(z)  = z\mod P(B, G)$ as per Equation \eqref{eq:circular-gran}.

For example, the multiple-order-up granularity $C_{\text{uinal},\text{katun}}$ for the Mayan calendar could be obtained using
\begin{align*}
C_{\text{uinal}, \text{baktun}}(z)  &=  C_{\text{uinal}, \text{tun}}(z) + P(\text{uinal}, \text{tun})C_{\text{tun},\text{katun}}(z) + P(\text{uinal},\text{katun})C_{\text{katun}, \text{baktun}}(z) \\
         &=  C_{\text{uinal}, \text{tun}}(z) + 18 \times{ C_{\text{tun},\text{katun}}(z)} + 18 \times 20 \times{C_{\text{katun}, \text{baktun}}(z)}
\end{align*}

, where $z$ is the index of the bottom granularity $kin$.
